Estimating duration using survival analysis

### 2023-10-20 cis-perfect-testing.tex
consistent notation for distributions
Incorporate Dani's comments

The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
However, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, model assumptions determined the size of the estimated tail estimated.
In this chapter, I develop a method to estimate the duration distribution using CIS data. CIS sampling is less frequent, but with a large sample size and long follow-up.

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
Flat priors are unintentionally very informative.
Hence, at least weakly informative priors should be used.
Of particular interest is a strongly informative prior to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
I conduct a simulation study to demonstrate this method's performance.
It succesfully recovers the truth when using reasonable priors (see \cref{perf-test:sec:simulation-study}).
In reality, the method must account for false negatives, discussed in \cref{E-imperf-test}.

@@

In the CIS each individualis tested  infrequently, either weekly or monthly (see\todo{refer to section with CIS intro}).
The infrequent testing complicates analysis of the data in two ways.
First, infection episodes can go undetected.
Second, the infection episode start and end times are known imprecisely.
This section explains these issues. The rest of the chapter develops a method to tackle  them.

The first complication in the analysis is that infection episodes can be undetected.
This means there are no positive tests associated with them.
Undetected episodes occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
For undetected episodes, there is no information on them episode; even the number that occurred is unknown.
Undetected episodes are important because they are shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
CIS has many undetected infections. This is because frequency of testing (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

@@

Uninformative Beta priors on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \Beta(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this can be drawn from the fact that these types of priors have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption[Uninformative priors for the hazard]{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

>>>>

When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
Duration distributions tend to be right-skewed.
Therefore, estiamtes of the mean are sensitive to the tail of the duration distribution.
In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
However, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, model assumptions drive the size of the estimated tails.
In this chapter, I develop a method to estimate the duration distribution using CIS data.
The CIS sampling is less frequent, but with a large sample size and long follow-up.

I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
I use a simulation study (see \cref{perf-test:sec:simulation-study}) to evaluate a range of priors for the survival distribution (described in \cref{perf-test:sec:parameters-priors}).
These include a prior to combine the information in the CIS with the results of \cref{E-ATACCC}.
Reasonable priors successfully recover the true distribution.
However, the method must be able to account for false negatives, discussed in \cref{E-imperf-test}.

@@

The CIS tests each individual infrequently, either weekly or monthly (see\todo{refer to section with CIS intro}).
The infrequent testing complicates analysis of the data in two ways.
First, infection episodes can go undetected.
Second, the infection episode start and end times are known imprecisely.
This section explains these issues. The rest of the chapter develops a method to tackle  them.

\subsection{Undetected infections}

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are important because they are shorter than detected episodes.
To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
The infection episode depicted begins at time 33 and has a 10-day duration.
If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
CIS has many undetected infections.
This is because interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days, see\todo{ref ATACCC chapter}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

\subsection{Double interval censoring}

@@

\subsection{Approach}

@@

Uninformative Beta priors on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this can be drawn from the fact that these types of priors have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption[Uninformative priors for the hazard]{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}
